A tutorial on spatiotemporal partially observed Markov process models via the R package spatPomp

### 2024-10-25 ms.Rnw
a few more edits

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{bbm} % for \mathbbm{1}
\graphicspath{{figure/}}

@@

\subsection{Inference algorithms inherited from pomp}
\label{sec:otherinference}
Objects of \class{spatPomp} inherit methods for inference from \class{pomp} objects implemented in the \pkg{pomp} package.
As discussed earlier, the IF2 algorithm \citep{ionides15} has been used for maximum likelihood parameter estimation in numerous applications.
IF2 can be used to check the capabilities of newer and more scalable inference methods on smaller examples for which it is known to be effective.
Extensions for Bayesian inference of the currently implemented high-dimensional particle filter methods (GIRF, ABF, EnKF, BPF) are not yet available.
Bayesian inference is available in \pkg{spatPomp} using the approximate Bayesian computing (ABC) method inherited from \pkg{pomp}, \code{abc()}.
ABC has previously been used for spatiotemporal inference \citep{brown2018approximate} and can also serve as a baseline method.
However, ABC is a feature-based method that may lose substantial information compared to full-information methods that work with the likelihood function rather than a summary statistic \citep{fasiolo16}.

@@

\code{ig1} is an object of \class{igirfd\_spatpomp} which inherits from \class{girfd\_spatpomp}.
A useful diagnostic of the parameter search is a plot of the change of the parameter estimates during the course of an \code{igirf()} run.
Each iteration within an \code{igirf} run provides a parameter estimate and a likelihood evaluation at that estimate.
The \code{plot} method for a \class{igirfd\_spatPomp} object shows the convergence record of parameter estimates and their likelihood evaluations.
As shown in \cref{fig:bm_igirf_convergence2}, this \code{igirf} search has allowed us to explore the parameter space and climb significantly up the likelihood surface to within a small neighborhood of the maximum likelihood.
The search took \Sexpr{myround(attr(ig1,"system.time")["elapsed"]/60,1)} minutes on one CPU core for this example with \Sexpr{length(unit_names(bm10))} spatial units.
Investigation of larger models may require multiple searches of the parameter space, started at various points, implemented using parallel runs of \code{igirf()}.

@@

<<bm_rho_prof_eval,eval=T,cache=F,echo=F>>=
rho_prof <- bake(file = paste0(bm_files_dir,"rho-profile.rds"),{
<<bm_rho_prof_code>>
attr(rho_prof,"cores") <- cores
rho_prof
})
@
Above, calling \code{igirf} on \code{ig1} imports all the previous algorithmic settings except for those that we explicitly modify.
Each row of \code{rho\_prof} now contains a parameter estimate its log-likelihood, with $\rho$ values fixed along a grid.
The MCAP 95\% confidence interval constructed by \code{mcap} uses \code{loess} to obtain a smoothed estimate of the profile likelihood function and then determines a confidence interval using by a cutoff based on the delta method applied to a local quadratic regression.
This cutoff is typically slightly larger than the asymptotic 1.92 cutoff for a standard profile likelihood confidence interval constructed assuming error-free likelihood maximization and evaluation.
<<bm_rho_prof_mcap>>=
rho_mcap <- mcap(rho_prof[,"loglik"],parameter=rho_prof[,"rho"])
rho_mcap$ci
@
Note that the data in \code{bm10} are generated from a model with $\rho=0.4$.
Even for this toy example, the profile likelihood confidence interval requires a non-trivial computational effort.
This calculation took $\Sexpr{myround(attr(rho_prof,"system.time")["elapsed"]/60/60,1)}$ hours using $\Sexpr{attr(rho_prof,"cores")}$ cores.
As a side-effect, the many searches involved in obtaining a profile likelihood ensure that the likelihood surface is extensively explored.

@@

The measles model divides the population of each city into susceptible, $S$, exposed, $E$, infectious, $I$, and recovered/removed, $R$, compartments.
The number of individuals in each compartment for city $\unit$ at time $t$ are denoted by $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
The latent state is $\vec{X}(t)=\big(X_1(t),\dots,X_{\Unit}(t)\big)$ with $X_{\unit}(t) = \big(S_{\unit}(t), E_{\unit}(t),I_{\unit}(t),R_{\unit}(t)\big)$.
The dynamics of the latent state can be written in terms of flows between compartments, together with flows into and out of the system, as follows:
\[
\left.
\begin{array}{lllllll}
\displaystyle dS_\unit(t) &=& dN_{BS,\unit}(t) &-& dN_{SE,\unit}(t) &-& dN_{SD,\unit}(t) \\
\displaystyle dE_\unit(t) &=& dN_{SE,\unit}(t) &-& dN_{EI,\unit}(t) &-& dN_{ED,\unit}(t) \\
\displaystyle dI_\unit(t) &=& dN_{EI,\unit}(t) &-& dN_{IR,\unit}(t) &-& dN_{ID,\unit}(t)
\end{array}\qquad\right\}\quad \mbox{for $\unit=1,\dots, \Unit$.}
\]
Here, $N_{SE,\unit}(t)$, $N_{EI,\unit}(t)$, and $N_{IR,\unit}(t)$ are counting process corresponding to the cumulative number of individuals transitioning between the compartments identified by the subscripts.
The recruitment of susceptible individuals into city $\unit$ is denoted by the counting process $N_{BS,\unit}(t)$, primarily modeling births.
Each compartment also has an outflow, written as a transition to $D$, primarily representing death, which occurs at a constant per-capita rate $\mu$.
The number of recovered individuals $R_\unit(t)$ in city $\unit$ is defined implicitly from $P_{\unit}(t)=S_{\unit}(t)+E_{\unit}(t)+I_{\unit}(t)+R_{\unit}(t)$.
$R_{\unit}(t)$ plays no direct role in the dynamics, beyond accounting for individuals not in any of the other classes.

@@

We construction the model described in \cref{sec:measles-math-model} for the simplified situation where $\alpha_{\unit}=1$, $\measlesImmigration_{\unit}=0$ and $\cohort_{\unit}=0$.
All other parameters have shared value across units, except for the initial value parameters.
A complete \pkg{spatPomp} representation of the model is provided in the source code for \code{he10()}.

@@

It is convenient (but not necessary) to supply a parameter vector of scientific interest for testing the model.
Here, we use a parameter vector with duration of infection and latent period both set equal to one week, following \citet{xia04}, and the basic reproduction number set to $\Rzero{ } =30$.
The gravitational constant, $g=1500$, was picked by qualitative visual matching of simulations.
<<meales_sim_par,echo=T,eval=T>>=
IVPs <- rep(c(0.032,0.00005,0.00004,0.96791),each=6) 
names(IVPs) <- paste0(rep(c('S','E','I','R'),each=6),1:6,"_0")
measles_params <- c(R0=30,A=0.5,muEI=52,muIR=52,muD=0.02,
  alpha=1,sigmaSE=0.01,rho=0.5,psi=0.1,g=1500,IVPs)
@
Special treatment is afforded to latent states that track accumulations of other latent states between observation times.
These accumulator variables should be reset to zero at each observation time.
The \code{unit\_accumvars} argument provides a facility to specify the unit-level names of accumulator variables, extending the \code{accumvars} argument to \code{pomp()}.
Here, there is one accumulator variable, \code{C}, which is needed since each case report corresponds to new reported infections accumulated over a measurement interval.
The pieces of the SpatPOMP are now added to \code{measles6} via a call to \code{\spatPomp}:
<<measles_full_spatpomp, echo=T, eval=T,cache=F>>=
measles6 <- spatPomp(
  data = measles6,
  covar = measles_covar,
  unit_statenames = c('S','E','I','R','C'),
  unit_accumvars = c('C'),
  paramnames = names(measles_params),
  rinit = measles_rinit,
  rprocess = euler(measles_rprocess, delta.t=1/365),
  dunit_measure = measles_dunit_measure,
  eunit_measure = measles_eunit_measure,
  vunit_measure = measles_vunit_measure,
  runit_measure = measles_runit_measure,
  dmeasure = measles_dmeasure,
  rmeasure = measles_rmeasure,
  globals = measles_globals
)
@
Here, we have not filled the \code{skeleton} and \code{munit\_measure} arguments, used by \code{girf} and \code{abfir}.
These can be found in the \pkg{spatPomp} package source code for \code{measles()}.

>>>>

\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{bbm} % for \mathbbm{1}
\usepackage{textcomp}
\graphicspath{{figure/}}

@@

\subsection{Inference algorithms inherited from pomp}
\label{sec:otherinference}
Objects of \class{spatPomp} inherit methods for inference from \class{pomp} objects implemented in the \pkg{pomp} package.
As discussed earlier, the IF2 algorithm \citep{ionides15} has been used for maximum likelihood parameter estimation in numerous applications.
IF2 can be used to check the capabilities of newer and more scalable inference methods on smaller examples for which it is known to be effective.
Extensions for Bayesian inference of the currently implemented high-dimensional particle filter methods (GIRF, ABF, EnKF, BPF) are not yet available.
Bayesian inference is available in \pkg{spatPomp} using the approximate Bayesian computing (ABC) method inherited from \pkg{pomp}, \code{abc()}.
ABC has previously been used for spatiotemporal inference \citep{brown2018approximate} and can also serve as a baseline method.
A related approach is the synthetic likelihood method of \cite{wood10} which can be implemented using \code{probe.match()}.
However, ABC and synthetic likelihood are feature-based methods that may lose substantial information compared to full-information methods that work with the likelihood function rather than a summary statistic \citep{fasiolo16}.

@@

\code{ig1} is an object of \class{igirfd\_spatpomp} which inherits from \class{girfd\_spatpomp}.
A useful diagnostic of the parameter search is a plot of the change of the parameter estimates during the course of an \code{igirf()} run.
Each iteration within an \code{igirf} run provides a parameter estimate and a likelihood evaluation at that estimate.
The convergence record of parameter estimates and their likelihood evaluations is stored in the \code{traces} slot of the resulting \class{igirfd\_spatPomp} object. 
The \code{plot} method for this class gives a graphical representation of the convergence traces, as shown in \cref{fig:bm_igirf_convergence2}.
We see that this \code{igirf} search has allowed us to explore the parameter space and climb significantly up the likelihood surface to within a small neighborhood of the maximum likelihood.
The search took \Sexpr{myround(attr(ig1,"system.time")["elapsed"]/60,1)} minutes on one CPU core for this example with \Sexpr{length(unit_names(bm10))} spatial units.
Investigation of larger models may require multiple searches of the parameter space, started at various points, implemented using parallel runs of \code{igirf()}.

@@

<<bm_rho_prof_eval,eval=T,cache=F,echo=F>>=
rho_prof <- bake(file = paste0(bm_files_dir,"rho-profile.rds"),{
<<bm_rho_prof_code>>
attr(rho_prof,"cores") <- cores
rho_prof
})
@
Above, calling \code{igirf} on \code{ig1} imports all the previous algorithmic settings except for those that we explicitly modify.
Each row of \code{rho\_prof} now contains a parameter estimate and its log-likelihood, with $\rho$ values fixed along a grid.
The MCAP 95\% confidence interval constructed by \code{mcap} uses \code{loess} to obtain a smoothed estimate of the profile likelihood function and then determines a confidence interval using by a cutoff based on the delta method applied to a local quadratic regression.
This cutoff is typically slightly larger than the asymptotic 1.92 cutoff for a standard profile likelihood confidence interval constructed assuming error-free likelihood maximization and evaluation.
<<bm_rho_prof_mcap>>=
rho_mcap <- mcap(rho_prof[,"loglik"],parameter=rho_prof[,"rho"])
rho_mcap$ci
@
Note that the data in \code{bm10} are generated from a model with $\rho=0.4$.
Even for this toy example, the profile likelihood confidence interval requires a non-trivial computational effort.
This calculation took $\Sexpr{myround(attr(rho_prof,"system.time")["elapsed"]/60/60,1)}$ hours using $\Sexpr{attr(rho_prof,"cores")}$ cores.
As a side-effect, the many searches involved in obtaining a profile likelihood ensure that the likelihood surface is extensively explored.

@@

The measles model divides the population of each city into susceptible, $S$, exposed, $E$, infectious, $I$, and recovered/removed, $R$, compartments.
The number of individuals in each compartment for city $\unit$ at time $t$ are denoted by $S_\unit(t)$, $E_\unit(t)$, $I_\unit(t)$, and $R_\unit(t)$.
The latent state is $\vec{X}(t)=\big(X_1(t),\dots,X_{\Unit}(t)\big)$ with $X_{\unit}(t) = \big(S_{\unit}(t), E_{\unit}(t),I_{\unit}(t),R_{\unit}(t)\big)$.
The dynamics of the latent state can be written in terms of flows between compartments, together with flows into and out of the system, as follows:
\[
\left.
\begin{array}{lllllll}
\displaystyle dS_\unit(t) &=& dN_{BS,\unit}(t) &-& dN_{SE,\unit}(t) &-& dN_{SD,\unit}(t) \\
\displaystyle dE_\unit(t) &=& dN_{SE,\unit}(t) &-& dN_{EI,\unit}(t) &-& dN_{ED,\unit}(t) \\
\displaystyle dI_\unit(t) &=& dN_{EI,\unit}(t) &-& dN_{IR,\unit}(t) &-& dN_{ID,\unit}(t)
\end{array}\qquad\right\}\quad \mbox{for $\unit=1,\dots, \Unit$.}
\]
Here, $N_{SE,\unit}(t)$, $N_{EI,\unit}(t)$, and $N_{IR,\unit}(t)$ are counting processes corresponding to the cumulative number of individuals transitioning between the compartments identified by the subscripts.
The recruitment of susceptible individuals into city $\unit$ is denoted by the counting process $N_{BS,\unit}(t)$, primarily modeling births.
Each compartment also has an outflow, written as a transition to $D$, primarily representing death, which occurs at a constant per-capita rate $\mu$.
The number of recovered individuals $R_\unit(t)$ in city $\unit$ is defined implicitly from $P_{\unit}(t)=S_{\unit}(t)+E_{\unit}(t)+I_{\unit}(t)+R_{\unit}(t)$.
$R_{\unit}(t)$ plays no direct role in the dynamics, beyond accounting for individuals not in any of the other classes.

@@

We construct the model described in \cref{sec:measles-math-model} for the simplified situation where $\alpha_{\unit}=1$, $\measlesImmigration_{\unit}=0$ and $\cohort_{\unit}=0$.
All other parameters have shared value across units, except for the initial value parameters.
A complete \pkg{spatPomp} representation of the model is provided in the source code for \code{he10()}.

@@

It is convenient (but not necessary) to supply a parameter vector for testing the model.
Here, we use a parameter vector with duration of infection and latent period both set equal to one week, following \citet{xia04}, and the basic reproduction number set to $\Rzero{ } =30$.
The gravitational constant, $g=1500$, was picked by qualitative visual matching of simulations.
<<meales_sim_par,echo=T,eval=T>>=
IVPs <- rep(c(0.032,0.00005,0.00004,0.96791),each=6) 
names(IVPs) <- paste0(rep(c('S','E','I','R'),each=6),1:6,"_0")
measles_params <- c(R0=30,A=0.5,muEI=52,muIR=52,muD=0.02,
  alpha=1,sigmaSE=0.01,rho=0.5,psi=0.1,g=1500,IVPs)
@
Special treatment is afforded to latent states that track accumulations of other latent states between observation times.
These accumulator variables should be reset to zero at each observation time.
The \code{unit\_accumvars} argument provides a facility to specify the unit-level names of accumulator variables, extending the \code{accumvars} argument to \code{pomp()}.
Here, there is one accumulator variable, \code{C}, which is needed since each case report corresponds to new reported infections accumulated over a measurement interval.
The pieces of the SpatPOMP are now added to \code{measles6} via a call to \code{\spatPomp}:
<<measles_full_spatpomp, echo=T, eval=T,cache=F>>=
measles6 <- spatPomp(
  data = measles6,
  covar = measles_covar,
  unit_statenames = c('S','E','I','R','C'),
  unit_accumvars = c('C'),
  paramnames = names(measles_params),
  rinit = measles_rinit,
  rprocess = euler(measles_rprocess, delta.t=1/365),
  dunit_measure = measles_dunit_measure,
  eunit_measure = measles_eunit_measure,
  vunit_measure = measles_vunit_measure,
  runit_measure = measles_runit_measure,
  dmeasure = measles_dmeasure,
  rmeasure = measles_rmeasure,
  globals = measles_globals
)
@
Here, we have not filled the \code{skeleton} and \code{munit\_measure} arguments, used by \code{girf} and \code{abfir}.
These can be found in the \pkg{spatPomp} package source code for \code{measles()}.
